Dialect: A Compact Line Notation Based on SMILES

### 2022-03-20 manuscript.tex
Avoids normative SMILES. closes #85

\hyphenation{organo-metallics}
\newcolumntype{L}{>{\arraybackslash}m{3cm}}
\lstset{basicstyle=\ttfamily,breaklines=true}

@@

\section*{SMILES Compatibility}

To maximize compatibility with contemporary software, Dialect is based on the SMILES language. Given the widespread adoption of "SMILES," arriving at a specific definition of the language is surprisingly challenging. No single document completely specifies SMILES syntax and semantics. The first description of the SMILES language in the primary literature is contained within a 1988 article (referenced here as "the article").\cite{weininger:1988} The article, by its own admission, is incomplete: "... isomeric SMILES [defined in the book chapter as covering isotopism, conformation, and configuration (p. 91)] is not otherwise covered in this paper." (p. 34). Weininger later refined and expanded SMILES in a 2003 book chapter,\cite{weininger:2008} referred to here as "the book chapter." A third major source of information is the Daylight Theory Manual (referred to here as "the manual"), a website operated by Daylight Chemical Information Systems, Inc.\cite{daylightTheory} Information about SMILES is also contained in various software packages and associated documentation, slide decks, websites, mailing lists, and private communications.

For the purpose of providing a foundation for Dialect, "SMILES" was defined as the language described in the article and the book chapter. All other sources --- irrespective of quality or apparent claim to authority --- were considered out of scope. The manual was disregarded because it merely recapitulated material in the article and chapter, and due to the possibility of future alteration. The remaining sources were disregarded due to lack of authority, poor accessibility, or both. This narrow focus simplifies the task of evaluating compatibility between Dialect and SMILES. However, it does not provide a method for testing compatibility due to the lack of an authoritative SMILES compliance suite.

It is nevertheless possible to compare Dialect and SMILES on a feature-by-feature basis. Dialect is intended to be a subset of SMILES, meaning that every feature present in Dialect is ideally also present in SMILES. As will be seen, certain factors make this undesirable in the strictest sense. With a few notable exceptions, however, Dialect can be viewed as a subset of SMILES.

The book chapter defines a fixed set of element symbols which even at the time of publication was obsolete (p. 83). The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha. To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og. Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC. These additions are one of the reasons Dialect can't be considered a strict subset of SMILES.

The article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules..." (p. 33). The comma character is explicitly disallowed in Dialect.

The book chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, results in multiple branching (p. 86). Dialect disallows the multiple branching constructs, including for example "*((*))*" and "*(((*)))*".

"Reaction SMILES" is an extension described in the book chapter (p. 89), but disallowed in Dialect. Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character. This precludes strings such as "*>>*" in Dialect. Furthermore, Dialect lacks the "map" attribute, used for atom-atom mapping in reactions, and its accompanying syntax.

Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@". The book chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@" (p. 94). The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1". None of these are supported by Dialect. Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter. Although some of these forms of configuration are supported today, it's unusual to find them all supported. The reason is simple: in the context in which SMILES is most often used, these forms of stereochemistry are rare. To support them places a burden on implementors with little payoff.

Along these lines, Dialect does not recognize atoms bearing either arsenic (As) or selenium (Se) as selectable. Doing so would increase the complexity of Dialect unnecessarily. Although the paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context." p35), default valences for the latter two atoms are not provided. Even if they were, the elements As and Se must be bracketed. As such, special rules for handling them in the context of pruning would be required. Moreover, molecules bearing the elements As or Se rarely exhibit DIME. Consider the arsenic analog of pyrrole ("arsole"). Selecting the arsenic-bearing atom is gratuitous because it does not contribute to DIME. Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing germanium (Ge), tin (Sn), or other heavy main group elements that may be discovered in the future to participate in DIME.

It's not clear whether SMILES allows a virtual hydrogen count on hydrogen itself. The article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence: "There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H]." (p. 34) If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as `[HH]`. The book chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g. [H][H], molecular hydrogen" (p. 97). To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself. For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]". Depending on interpretation, this support may or may not render Dialect a superset of SMILES.

The book chapter ascribes ambiguous meaning to detachments: "... In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other." (p. 88). The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.

SMILES assigns no explicit upper or lower bounds to numerical atomic attributes. These boundaries are nevertheless crucial for implementors, who often seek a data type just large enough to prevent underflow or overflow. For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and virtual hydrogen (\ttt{hydrogens}, 0 $\leq$ value \textless 10). Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be obvious in SMILES, the lack of specificity forces implementors to complete work that should have been done by the language's creator.

Dialect provides a few semantic clarifications not fully addressed in SMILES. A detailed algorithm for determining implicit hydrogen count is provided, together with the required valence table. Unlike SMILES, Dialect explicitly considers the case of computing implicit hydrogens on selected atoms. Dialect also provides detailed algorithms, absent in SMILES, for selection and deselection. These are based in graph theory rather than the ambiguous and overloaded chemical concept of "aromaticity" used in SMILES. Uniquely, Dialect introduces the concepts of "pruning" and "gratuitous selection." Neither the paper nor the book chapter indicate those bonds that can be promoted during deselection, but Dialect does. Finally, Dialect syntax is based on a formal grammar and tooling rooted in decades of computer science research, whereas SMILES syntax is based for the most part on imprecise natural language descriptions.

Dialect also resolves some internal contradictions within SMILES. The article reports 3 as the only default valence for nitrogen (p. 34), whereas the book chapter reports two default valences: 3 and 5 (p. 84). Dialect uses the latter default valences. Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the article as: B; C; N; O; P; S; F; Cl; Br; and I (p. 32). However, an example later in the article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string (p. 35). Dialect explicitly rejects strings containing unbracketed hydrogen atoms. On the topic of atom selection ("aromaticity"), the article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion." (p. 34). The book chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc", p. 85). Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.

\begin{table}
    \caption{Some differences between SMILES and Dialect.}
    \centering
    \begin{tabular}{p{0.55\linewidth} p{0.15\linewidth} p{0.15\linewidth}}
        \hline
        Feature & SMILES & Dialect \\
        \hline
        element symbol Ha & yes & no \\
        element symbols Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og & no & yes \\
        future element symbols approved by IUPAC & no & yes \\
        comma symbol (\ttt{,}) & maybe & no \\
        multiple branching such as "*((*))*" & yes & no \\
        reactions using greater than symbol (\ttt{>}) & yes & no \\
        atomic "map" attribute & yes & no \\
        extended stereodescriptors & yes & no \\
        use of stereodescriptors on odd cumulene centers & yes & no \\
        virtual hydrogen count on hydrogen & no & yes \\
        detachments are bonds of "formal order zero" & maybe & no \\
        upper and lower bounds on atomic properties & no & yes \\
        nitrogen default valence includes 5 & maybe & yes \\
        unbracketed hydrogen atom & maybe & no \\
        acyclic atom selection & maybe & yes \\
        atoms with elements As and Se are selectable & yes & no \\
        non-elided bonds may be promoted during selection & maybe & no \\
        \hline
    \end{tabular}
    \label{table:feature-comparison}
\end{table}

>>>>

\hyphenation{organo-metallics Proto-SMILES}
\newcolumntype{L}{>{\arraybackslash}m{3cm}}
\lstset{basicstyle=\ttfamily,breaklines=true}

@@

\section*{Compatibility}

For maximum compatibility with existing software, Dialect was designed as a \textit{language subset} of SMILES. A language subset contains all of the syntax and semantics of its parent language, but adds none of its own. In principle this means that a feature present in Dialect should also be present in SMILES. Conversely, a feature present in SMILES may or may not also be present in Dialect.

Language subsets have ample precedent in computer science. A well-known example is MISRA C,\cite{misraC} a subset of the C programming language that aims to eliminate "known undefined or otherwise dangerous behavior."\cite{hatton:2007} Another example is the subset of JavaScript described by Crockford whose purpose is to "chip away at the features that are not beautiful until the language's true nature reveals itself."\cite{crockford:2008} Language subsets arise for two main reasons. First, designing a language presents many opportunities for errors small and large. Second, popular languages inevitably bring with them many users with diverse problems that demand new features. Some of these features cause future problems of their own. A language subset can improve both situations by eliminating the most problematic features.

Although similar considerations apply to SMILES, its case is different in one important way: SMILES was never publicly described with high precision. Overviews of the language have been published by both Weininger and Daylight. But the kind of technical documentation suitable for software development was never made public. Both the OpenSMILES and SMILES+ initiatives cite this shortfall as motivation.

Nevertheless, several SMILES software implementations have been released as noted previously. Each one represents its own unique set of decisions about how to reduce the published broad SMILES descriptions to a specification detailed enough for software development. Some implementations have themselves been reverse-engineered for clues. The Daylight implementation in particular is sometimes invoked as a de facto SMILES specification. Unfortunately, access to this implementation is restricted by commercial licensing, as is the case with several SMILES implementations. Moreover, reverse engineering is unlikely to address every question because of the vast search space embodied by SMILES. Although some implementations are released under Open Source licenses, they are neither authoritative nor unanimous in their interpretations.

A publicly-available and precise subset of SMILES would offer a new way to address the problem. If sufficiently constrained, the subset should be interpretable by all existing software without modification. Moreover, most strings written by existing software would be interpretable by software based on the language subset. Because its description would be both public and precise, the language subset could serve as the basis for many implementations with strong consensus around both syntax and semantics.

Unfortunately, the choice of the Dialect base language is fraught with difficulty on technical, historical, and cultural grounds. Simply put, "SMILES" has come to mean different things to different people. To some, SMILES is the language defined in public documentation. To others, SMILES is the Daylight implementation source code. Some may view SMILES as the collective definitions contained in both published documents and all software. To others, SMILES isn't even definable.

These problems can be avoided by considering as a base language not SMILES itself, but a \textit{protolanguage}. A protolanguage is a hypothetical language from which a family of languages can be derived. A suitable SMILES protolanguage would ideally contain all of the features consistent with the many ways in which the term "SMILES" is used today.

Accordingly, Dialect's base language is "ProtoSMILES," a SMILES protolanguage. ProtoSMILES was developed through a process of elimination that started with every SMILES document ever published and every SMILES software implementation, Open Source or not. The field was narrowed through two limitations: (1) the source must be authoritative; and (2) the source must be publicly accessible. In this context, authoritative means that the source was created by either Weininger himself or Daylight. They alone can speak most authoritatively about what SMILES is. The requirement for public accessibility ensures that, even if inconvenient, claims of compatibility can be tested. The following documents were identified as candidate ProtoSMILES sources:

\begin{itemize}
    \item Weininger's report to the US Environmental Protection Agency (herein "The Article").\cite{anderson:1987}
    \item Weininger's original publication (herein "The Article").\cite{weininger:1988}
    \item Weininger's 2003 book chapter, revised in 2008 (herein "The Book Chapter").\cite{weininger:2008}
    \item An online manual maintained by Daylight (herein "The Manual").
    \item The Daylight's toolkit, which implements a SMILES reader and writer (herein "The Toolkit").\cite{daylightToolkit}
\end{itemize}

The Toolkit was eliminated for three reasons. First, its source code is not available, forcing an indirect reverse-engineering approach to language definition. Second, as a product of the company Weininger started, the Toolkit is likely to recapitulate the other sources. Third, as a commercial product without source code, the Toolkit may not always be available in the future.

The Report was eliminated mainly because the contents of this older document contradict the three remaining sources. For example, The Report contains no mention of bracket syntax, using markedly different notation instead. The report also supports unusual characters including the exclamation mark (\ttt{!}). Semantically, The Report supports "special states" for nitrogen-containing atoms.

The Manual was disregarded as a ProtoSMILES source because it merely recapitulates content already present in The Paper and The Book Chapter. Removing this source simplifies the construction of ProtoSMILES without changing its definition. A secondary consideration was the possibility that updates to this web page could change the meaning of ProtoSMILES. However web archives indicate that neither styling nor content of the Manual have changed in at least ten years.

The only remaining sources are The Paper and the Book Chapter. These are the two works that collectively define ProtoSMILES. The remainder of this section compares and contrasts ProtoSMILES with Dialect. The main difficulty with this approach is that the works defining ProtoSMILES not only contradict each other in some ways, but they also contradict themselves internally. These contradictions will be noted when meaningful.

The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete.\cite[p. 83]{weininger:2008} The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha. To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og. Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC. The addition of these element symbols means that Dialect can't be considered a strict subset of SMILES. The alternative would, however, leave dialect without the ability to express every kind of IUPAC-recognized element.

The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...".\cite[p. 33]{weininger:1988} The comma character is explicitly disallowed in Dialect.

The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses.\cite[p. 86]{weininger:2008} Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".

"Reaction SMILES" is an extension described in the Book Chapter,\cite[p. 89]{weininger:2008} but disallowed in Dialect. Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character. This precludes strings such as "*>>*" in Dialect. Furthermore, Dialect lacks the "map" attribute, used for atom-atom mapping in reactions, and its accompanying syntax.

Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@". The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@".\cite[p. 94]{weininger:2008} The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1". None of these are supported by Dialect. Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter. Although some of these forms of configuration are supported today, it's unusual to find them all supported. The reason is simple: in the context in which Dialect will most often used, these forms of stereochemistry are rare. To support them burdens implementors with little payoff.

ProtoSMILES supports the concept of "aromatic bond" (":"), but Dialect does not. Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided. The Book Chapter notes "Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond). "-" (single) and ":" (aromatic) bond symbols may always be omitted on input".\cite[p 85]{weininger:2008} This passage implies that aromatic bonds serve no purpose. Accordingly, Dialect does not allow them and the syntax does not include the colon symbol (\ttt{:}).

Along these lines, Dialect does not recognize atoms bearing either arsenic (As) or selenium (Se) as selectable. Doing so would increase the complexity of the language unnecessarily. Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context."),\cite[p. 35]{weininger:1988} default valences for the latter two atoms are not provided. Even if they were, the elements As and Se must be bracketed. As such, special rules for handling them in the context of pruning would be required. Moreover, molecules bearing the elements As or Se are often made "aromatic" unnecessarily in SMILES. Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing silicon, germanium, tin, or other heavy main group elements that may be discovered in the future to participate in DIME.

It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself. The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].").\cite[p. 34]{weininger:1988} If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]". The Book Chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g. [H][H], molecular hydrogen".\cite[p 97]{weininger:2008} To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself. For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]". Depending on interpretation, this feature may or may not render Dialect a superset of SMILES.

ProtoSMILES ascribes ambiguous meaning to detachments. As noted in the Book Chapter, "... In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.".\cite[p 88]{weininger:2008} The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.

ProtoSMILES assigns no explicit upper or lower bounds to numerical atomic attributes. These boundaries are nevertheless crucial for implementors, who seek the data types necessary to prevent underflow and overflow. For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and hydrogens, 0 $\leq$ value \textless 10). Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior. ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].").\cite[p 32]{weininger:1988} However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.

Dialect provides a few semantic clarifications not fully addressed in ProtoSMILES. A detailed algorithm for determining implicit hydrogen count is provided, together with the required valence table. Unlike ProtoSMILES, Dialect explicitly considers the case of computing implicit hydrogens on selected atoms. Dialect also provides detailed algorithms, absent in SMILES, for selection and deselection. These are based in graph theory rather than the ambiguous and overloaded chemical concept of "aromaticity" used in ProtoSMILES. Uniquely, Dialect introduces the concepts of "pruning" and "gratuitous selection." ProtoSMILES does not specify those bonds that can be promoted during deselection, but Dialect does. ProtoSMILES does not restrict partial parity bonds to those atoms also possessing at least one double bond, whereas Dialect does. Finally, Dialect syntax is based on a formal grammar and tooling rooted in decades of computer science research, whereas ProtoSMILES syntax is based for the most part on imprecise natural language descriptions.

Dialect also resolves some internal contradictions within ProtoSMILES. The Article reports the only default valence for nitrogen as three (p. 34),\cite[p 34]{weininger:1988} whereas the Book Chapter reports two default valences: 3 and 5.\cite[p 84]{weininger:2008} Dialect uses the latter default valences. Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I.\cite[p 32]{weininger:2008} However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string.\cite[p 35]{weininger:1988} Dialect explicitly rejects strings containing unbracketed hydrogen atoms. On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."\cite[p 34]{weininger:1988} The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc").\cite[p 85]{weininger:2008} Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.

\begin{table}
\begin{minipage}{\textwidth}
\renewcommand*\footnoterule{}
    \caption{Some differences between ProtoSMILES and Dialect.}
    \centering
    \begin{tabular}{p{0.50\linewidth} p{0.20\linewidth} p{0.15\linewidth}}
        \hline
        Feature & ProtoSMILES\footnote{Values of "maybe" indicate ambiguous source descriptions.} & Dialect \\
        \hline
        element symbol Ha & yes & no \\
        element symbols Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og & no & yes \\
        future element symbols approved by IUPAC & no & yes \\
        comma symbol (\ttt{,}) & maybe & no \\
        multiple branching such as "*((*))*" & yes & no \\
        reactions using greater than symbol (\ttt{>}) & yes & no \\
        atomic "map" attribute & yes & no \\
        extended stereodescriptors (e.g., "@@@", "@OH1", etc.) & yes & no \\
        use of stereodescriptors on odd cumulene centers & yes & no \\
        virtual hydrogen count on hydrogen (e.g., "[HH]") & maybe & yes \\
        detachments are bonds of "formal order zero" & maybe & no \\
        upper and lower bounds on numerical atomic properties & no & yes \\
        nitrogen default valence includes 5 & maybe & yes \\
        unbracketed hydrogen atom & maybe & no \\
        acyclic atom selection & maybe & yes \\
        atoms with elements As and Se are selectable & yes & no \\
        non-elided bonds may be promoted during selection & maybe & no \\
        isolated partial parity bond & yes & no \\
        repeated charge notation (e.g., "++", "+++", etc.) & yes & no \\
        aromatic bond (":") & yes & no \\
        \hline
    \end{tabular}
    \label{table:feature-comparison}
\end{minipage}
\end{table}

With only one clear-cut exception, Table~\ref{table:feature-comparison} shows Dialect to be a subset of ProtoSMILES. The exceptional case deals with the validity certain transfermium element symbols already recognized by IUPAC and to be recognized in the future. Ambiguous exceptions include those for which the ProtoSMILES column feature support is uncertain ("maybe") but the Dialect column is affirmative ("yes"). If ProtoSMILES support tends toward "no," then support by Dialect would break out of the ProtoSMILES subset. There are three such cases. The first is dihydrogen represented using the string "[HH]". The second ambiguous case is whether or not the default valence of nitrogen includes the value five. The third case is whether or not ProtoSMILES recognizes acyclic "aromatic" atoms as valid.
